Understanding Factors Contributing to a Successful Himalayan Expedition

### 2024-04-19 paper/paper.qmd
Final changes

---
title: "Understanding Factors Contributing to a Successful Himalayan Expedition"
subtitle: "Analysis using data from Himalayan expeditions from 1905 through Spring 2019 leveraging Bayesian Logistic Regression"
author: Kaavya Kalani
thanks: "Code and data supporting this analysis is available at: [https://github.com/kaavyakalani26/himalayan-expeditions-analysis](https://github.com/kaavyakalani26/himalayan-expeditions-analysis)"
date: today
date-format: long
abstract: "This study analyses the relationship between various demographic, environmental and geographic factors and the probability of success in attempting to summit a peak. Data from Himalayan expeditions in Nepal from 1905 through Spring 2019 is used and a Bayesian Logistic Regression model is leveraged to analyse the trends and factors influencing a successful summit. Young age, being male and embarking on the expedition in spring or autumn are some factors that increase one's chances of a successful summit. The insights from this study aim to assist in future expedition planning, risk management, and safety protocols."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

# Introduction
Mountaineering, with its blend of adventure and challenge, has captivated explorers for generations. It represents a pinnacle of human endeavor, testing physical strength, mental fortitude, and strategic planning against some of the most formidable natural landscapes on Earth. Amidst the allure of conquering majestic peaks, a critical question arises: what factors contribute to the likelihood of a successful summit attempt? This paper delves into precisely this.

@@

  # Create a bar plot
  ggplot(data, aes_string(x = x)) +
    geom_bar(fill = "lightblue") +
    labs(x = x_label, y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

@@

@@

@@

# Plot for the predicted success probability by age and sex
ggplot(predictions, aes(x = age_range, y = success_probability, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "Age",
    y = "Success Probability"
  ) +
  facet_grid(sex ~ .) +
  theme_minimal()
```
@fig-age-sex shows the predicted success probability based on the age and sex of expedition members. It is evident that success probabilities are slightly higher for males across all ages, consistent with the findings in @fig-model-coefficients. However, when age is factored in, these differences vary. For individuals under 50, the success probabilities for males and females are similar, but for those aged 51 and above, the gap widens significantly. Success probabilities are notably high, exceeding 0.8 for younger individuals but decrease with age. The sharp decline in the 81-90 age category may be attributed to the smaller number of expedition members in that age group, resulting in less reliable outcomes.

@@

The physical demands of mountaineering, requiring exceptional strength and endurance, align with biological differences between sexes, as men typically have greater muscle mass and strength. This may explain why men are more drawn to mountaineering due to their physical advantages. To have the same physical strength, a woman would need to work much harder.

Despite the numerical difference, an interesting observation emerges from the analysis of success probabilities. @fig-model-coefficients and @fig-age-sex show how the success probabilities of men and women are not extremelyx different. While men have a slightly higher likelihood of reaching the summit, women show commendable success rates despite their lower representation in expeditions.

@@

## Spring and Autumn are the best time for expeditions
Autumn and spring, as transitional seasons, provide favorable conditions for mountaineering expeditions. The periods offer moderate temperatures and more stable weather patterns compared to the extremes of summer and winter. @fig-model-coefficients and @fig-height-seasons show the difference in probabilities of success by seasons and highlight highest chances of success in spring followed by autumn.

@@

In conclusion, autumn and spring are the optimal seasons for mountaineering expeditions, and season selection is a crucial factor in expedition planning.

@@

While age-related decline in mountaineering performance is noticeable, it is essential to recognize that success in mountaineering is not solely determined by physical capabilities. Experience, skill, preparation, and mindset also play crucial roles in summit success. Older climbers can compensate for physiological limitations through greater experience, strategic planning, and mental resilience, enabling them to pursue their mountaineering goals despite age-related challenges.

@@

Group dynamics play a crucial role in mountaineering, affecting decision-making, safety, resource allocation, and overall team morale. Studying how group size influences summit success can provide insights into optimal team composition, leadership structures, and communication strategies. It can also highlight the trade-offs between the benefits of larger teams and the challenges of coordinating and managing them effectively.

@@

pp_check(expeditions_model) +
  theme(legend.position = "bottom")
```

@@

## Markov chain Monte Carlo (MCMC) Convergence Check
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace1
#| fig-cap: "Trace plot"
#| fig-subcap: ["Intercept", "sexM", "soloTRUE", "seasonsSpring", "seasonsSummer", "seasonsWinter", "age_rangeUnder 18",  "age_range31-40", "age_range41-50", "age_range51-60", "age_range61-70", "age_range71-80", "age_range81-90"]
#| layout-ncol: 3

plot(expeditions_model, "trace", "(Intercept)")
plot(expeditions_model, "trace", "sexM")
plot(expeditions_model, "trace", "soloTRUE")
plot(expeditions_model, "trace", "seasonsSpring")
plot(expeditions_model, "trace", "seasonsSummer")
plot(expeditions_model, "trace", "seasonsWinter")
plot(expeditions_model, "trace", "age_rangeUnder 18")
plot(expeditions_model, "trace", "age_range31-40")
plot(expeditions_model, "trace", "age_range41-50")
plot(expeditions_model, "trace", "age_range51-60")
plot(expeditions_model, "trace", "age_range61-70")
plot(expeditions_model, "trace", "age_range71-80")
plot(expeditions_model, "trace", "age_range81-90")
```
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace2
#| fig-cap: "Trace plot"
#| fig-subcap: ["height_range5750 - 6099", "height_range6100 - 6449", "height_range6450 - 6799", "height_range6800 - 7149", "height_range7150 - 7499", "height_range7500 - 7849", "height_range7850 - 8199", "height_range8200 - 8549", "height_range8550 - 8900"]
#| layout-ncol: 3

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-rhat
#| fig-cap: "Rhat plot"

>>>>

---
title: "Understanding Factors Contributing to a Successful Himalayan Expedition"
subtitle: "Analysis using data from Himalayan expeditions from 1905 through Spring 2019 leveraging Bayesian Logistic Regression"
author: Kaavya Kalani
thanks: "Code and data supporting this analysis is available at: [https://github.com/kaavyakalani26/himalayan-expeditions-analysis](https://github.com/kaavyakalani26/himalayan-expeditions-analysis)"
date: today
date-format: long
abstract: "This study analyses the relationship between various demographic, environmental and geographic factors and the probability of success in attempting to summit a peak. Data from Himalayan expeditions in Nepal from 1905 through Spring 2019 is used and a Bayesian Logistic Regression model is utilised to analyse the trends and factors influencing a successful summit. Young age, being male and embarking on the expedition in spring or autumn are some factors that increase one's chances of a successful summit. The insights from this study aim to assist in future expedition planning, risk management, and safety protocols."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

# Introduction
Mountaineering, with its blend of adventure and challenge, has captivated explorers for generations. It represents a pinnacle of human endeavor, testing physical strength, mental fortitude, and strategic planning against some of the most formidable natural landscapes on Earth. Amidst the allure of conquering majestic peaks, a critical question arises: what factors contribute to the likelihood of a successful summit attempt? This paper tries to unravel precisely this.

@@

  # Custom labels for "sex" variable
  sex_labels <- c("Female", "Male")

  # Create a bar plot
  p <- ggplot(data, aes_string(x = x)) +
    geom_bar(fill = "lightblue") +
    labs(x = x_label, y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  # If x is "sex", set custom labels
  if (x == "sex") {
    p <- p + scale_x_discrete(labels = sex_labels)
  }

@@

  # Custom labels for "sex" variable
  sex_labels <- c("Female", "Male")

@@

  # If x is "sex", set custom labels
  if (x_var == "sex") {
    p <- p + scale_x_discrete(labels = sex_labels)
  }

@@

# Plot for the predicted success probability by age and sex
ggplot(predictions, aes(x = age_range, y = success_probability, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "Age",
    y = "Success Probability"
  ) +
  facet_grid(sex ~ .) +
  scale_fill_discrete(labels = c("Female", "Male")) +
  theme_minimal()
```
@fig-age-sex shows the predicted success probability based on the age and sex of expedition members. It is evident that success probabilities are slightly higher for males across all ages, consistent with the findings in @fig-model-coefficients. However, when age is factored in, these differences vary. For individuals under 50, the success probabilities for males and females are similar, but for those aged 51 and above, the gap widens significantly. Success probabilities are notably high, exceeding 0.8 for younger individuals but decrease with age. The sharp decline in the 81-90 age category may be attributed to the smaller number of expedition members in that age group, resulting in less reliable outcomes.

@@

The physical demands of mountaineering, requiring exceptional strength and endurance [@Strength], align with biological differences between sexes, as men typically have greater muscle mass and strength [@Sex]. This may explain why men are more drawn to mountaineering due to their physical advantages. To have the same physical strength, a woman would need to work much harder.

Despite the numerical difference, an interesting observation emerges from the analysis of success probabilities. @fig-model-coefficients and @fig-age-sex show how the success probabilities of men and women are not extremely different. While men have a slightly higher likelihood of reaching the summit, women show commendable success rates despite their lower representation in expeditions.

@@

## Spring and Autumn are the best time for expeditions
Autumn and spring, as transitional seasons, provide favorable conditions for mountaineering expeditions. The periods offer moderate temperatures and more stable weather patterns compared to the extremes of summer and winter. @fig-model-coefficients and @fig-height-seasons show the difference in probabilities of success by seasons and highlight higher chances of success in spring and autumn. @Season provides tips for summit success when climbing Mt. Everest and suggest that the best time to climb it is in the Autumn/ Fall season. This is in-line with our findings.

@@

In conclusion, autumn and spring are the optimal seasons for mountaineering expeditions, and season selection is an important factor in expedition planning.

@@

@Age highlight some of the same issues and factors when analysing endurance performance in masters athletes.

While age-related decline in mountaineering performance is noticeable, it is essential to recognize that success in mountaineering is not solely determined by physical capabilities. Experience, skill, preparation, and mindset also play an important roles in summit success. Older climbers can compensate for physiological limitations through greater experience, strategic planning, and mental resilience, enabling them to pursue their mountaineering goals despite age-related challenges.

@@

Group dynamics play an important role in mountaineering, affecting decision-making, safety, resource allocation, and overall team morale. Studying how group size influences summit success can provide insights into optimal team composition, leadership structures, and communication strategies. It can also highlight the trade-offs between the benefits of larger teams and the challenges of coordinating and managing them effectively.

@@

# Posterior check
pp_check(expeditions_model) +
  theme(legend.position = "bottom")
```

@@

## Markov chain Monte Carlo (MCMC) Convergence Check
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-trace1
#| fig-cap: "Trace plot"
#| fig-subcap: ["Intercept", "sexM", "soloTRUE", "seasonsSpring", "seasonsSummer", "seasonsWinter", "age_rangeUnder 18",  "age_range31-40", "age_range41-50", "age_range51-60", "age_range61-70", "age_range71-80", "age_range81-90"]
#| layout-ncol: 3

# Split plots as all at once is unreadable
plot(expeditions_model, "trace", "(Intercept)")
plot(expeditions_model, "trace", "sexM")
plot(expeditions_model, "trace", "soloTRUE")
plot(expeditions_model, "trace", "seasonsSpring")
plot(expeditions_model, "trace", "seasonsSummer")
plot(expeditions_model, "trace", "seasonsWinter")
plot(expeditions_model, "trace", "age_rangeUnder 18")
plot(expeditions_model, "trace", "age_range31-40")
plot(expeditions_model, "trace", "age_range41-50")
plot(expeditions_model, "trace", "age_range51-60")
plot(expeditions_model, "trace", "age_range61-70")
plot(expeditions_model, "trace", "age_range71-80")
plot(expeditions_model, "trace", "age_range81-90")
```
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-trace2
#| fig-cap: "Trace plot"
#| fig-subcap: ["height_range5750 - 6099", "height_range6100 - 6449", "height_range6450 - 6799", "height_range6800 - 7149", "height_range7150 - 7499", "height_range7500 - 7849", "height_range7850 - 8199", "height_range8200 - 8549", "height_range8550 - 8900"]
#| layout-ncol: 3

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-rhat
#| fig-cap: "Rhat plot"
